---
title: "Pollinator contribution to crop yield"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

::: {.content-visible when-format="html"}
# Results
:::

```{r load__more_things, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# source the data created using "scripts/read_data.R", which is triggered by track_estimates.R file.
load("scripts/temp_data.RData") 
library(lme4)

```

### 1) What are the relative contributions of honey bees versus wild insects to crop visitation worldwide?

We observe a large variation in the relative contribution of honey bee visitation rates and other wild insects within and across crops. While some crops are solely visited by wild insects in some areas, others are mainly visited by honeybees.

```{r Fig1, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# analysis: proportion of visits by HB vs WI

# fraction of visits by WI boxplot

# calculate fraction of "visits" by each group
field_data8a$fracWI = field_data8a$wild_insects/field_data8a$all_insects
field_data8a$fracHB = field_data8a$honeybee/field_data8a$all_insects

# reorder factor levels of study so they will plot in order of increasing median when we run a boxplot
field_data8a$study_id2 <- reorder(field_data8a$study_id2, field_data8a$fracWI, median, na.rm=T)

par(mar=c(4,4,4,4))
boxplot(data=field_data8a, fracWI ~ study_id2, col=c("light gray"), xaxt = 'n', xlab="study", ylab="proportion of visits by wild insects", las = 1)
axis(1,seq(5,90,5))

```

**Fig 1:** This figure shows the proportion of total visits provided by wild insects (vs. honey bee) for each pollination study. In these boxplots, the bold center line is the median, the hollow boxes cover the interquartile range (IQR), and the whiskers extend to the most extreme points within 1.5\*IQR from the median. If any points are more extreme than this, they are plotted as gray circles.

### 2) What are the relative contributions of honey bees versus wild insects to crop yield worldwide?

```{r model, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# analysis: mixed models for HB vs WI------

# for this we are using field_data8a, which has HB and WI, and where we have entered manual 0's for HB z-scores in 2 systems where they were =0.
# note that some of these studies are bees only, and many do not have richness data

# note: the following models will return a warning of "boundary (singular) fit: see ?isSingular".  I think this is due to using z-scores, which make the mean of all studies =0, so it should not be a problem for the models.

# analysis: mixed models for HB vs WI vs Richness------
# for this we are using field_data9a, which is the subset of studies providing data for HB, WI, and Richness

# I didn't get any convergence warnings when I ran these with the current data, but if it does happen, try adding one of these to the lmer statement:
# control = lmerControl(optimizer="Nelder_Mead")
# control = lmerControl(optimizer ='optimx', optCtrl=list(method='L-BFGS-B'))

# below I have coded the model with 1+ in the random effect syntax, however, it is worth considering whether we should use 0+ since all studies are z-scored and thus constrained to have a mean of 0. The logic would be that there is no need to try to estimate separate intercepts for each study. Regardless we would estimate the random slopes as usual.

mod_R18 = lmer(data=field_data9a, yield_z ~ wild_insects_z + honeybee_z + richness_z + wild_insects_z:honeybee_z + wild_insects_z:richness_z + honeybee_z:richness_z 
+ (1 + wild_insects_z|study_id2)  + (1 + honeybee_z|study_id2) + (1 + richness_z|study_id2) + (1 + wild_insects_z:honeybee_z|study_id2) + (1 + wild_insects_z:richness_z|study_id2) + (1 + honeybee_z:richness_z|study_id2), na.action=na.fail, REML=T)
#summary(mod_R18)

# note that the following calculations will only work with the 1+ random effect coding discussed above
a=fixef(mod_R18)
a_wi = a[2]		#overall slope of wild insects 
a_hb = a[3]		#overall slope of honey bee 
a_rich = a[4]
b=ranef(mod_R18)
qq <- attr(ranef(mod_R18, condVar = TRUE)[[1]], "postVar") # Extract the variances of the random effects
qq_wi = qq[[1]]
qq_hb = qq[[2]]
qq_rich = qq[[3]]
e_wi=(sqrt(qq_wi))	
e_wi=e_wi[2,2,]
e_hb=(sqrt(qq_hb))	
e_hb=e_hb[2,2,]
e_rich=(sqrt(qq_rich))	
e_rich=e_rich[2,2,]	
b_wi = b[[1]][2]	# slope of wild insects by crop?  lmer calls this the conditional mean
b_hb = b[[1]][4]	# slope of hb by crop?  lmer calls this the conditional mean
b_rich = b[[1]][6]	# slope of rich by crop?  lmer calls this the conditional mean
mean_wi=(b_wi+a_wi)			# mean ...  add by crop part to overall slope
ci_lower_wi=(b_wi+a_wi)-(e_wi*2)	# lower CI
ci_upper_wi=(b_wi+a_wi)+(e_wi*2)	# upper CI
mean_hb=(b_hb+a_hb)			# mean ...  add by crop part to overall slope
ci_lower_hb=(b_hb+a_hb)-(e_hb*2)	# lower CI
ci_upper_hb=(b_hb+a_hb)+(e_hb*2)	# upper CI
mean_rich=(b_rich+a_rich)			# mean ...  add by crop part to overall slope
ci_lower_rich=(b_rich+a_rich)-(e_rich*2)	# lower CI
ci_upper_rich=(b_rich+a_rich)+(e_rich*2)	# upper CI

###
modelout = cbind(mean_wi, ci_lower_wi, ci_upper_wi, mean_hb, ci_lower_hb, ci_upper_hb, mean_rich, ci_lower_rich, ci_upper_rich)
colnames(modelout)[1] = "mean_wi"
colnames(modelout)[2] = "ci_lower_wi"
colnames(modelout)[3] = "ci_upper_wi"
colnames(modelout)[4] = "mean_hb"
colnames(modelout)[5] = "ci_lower_hb"
colnames(modelout)[6] = "ci_upper_hb"
colnames(modelout)[7] = "mean_rich"
colnames(modelout)[8] = "ci_lower_rich"
colnames(modelout)[9] = "ci_upper_rich"
modelout$crop = rownames(modelout)
#modelout
modelout_mod_R18 = modelout

```

Similarly, we observe a large variation on the effect of pollinator richness, wild and honeybee visitation rates on crop yields, with an overall small, but positive effect size.

```{r Fig2, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# plot figure 2
# plots random effect estimates by study, similar to Garibaldi et al 2013 fig 2b -----

par(mfrow=c(1,3))

modelout_mod_R18 = modelout_mod_R18[order(modelout_mod_R18$mean_wi), , drop = FALSE]
plot(1:nrow(modelout_mod_R18), modelout_mod_R18$mean_wi, las = 1, pch=19, col="#00bb5c", ylim=c(-.75,.75), ylab="wild insect slope (mean and 95% CI)", cex=1.5, xaxt = "n", xlab="crop study", main="Wild Insects")
axis(1, at=seq(0,nrow(modelout_mod_R18),10), labels=seq(0,nrow(modelout_mod_R18),10))
arrows(1:nrow(modelout_mod_R18), modelout_mod_R18$ci_lower_wi, 1:nrow(modelout_mod_R18), modelout_mod_R18$ci_upper_wi, length=0.05, angle=90, code=3, col="#00bb5c", lwd=1.5)
abline(h=a_wi, lty="dashed", lwd=2)
abline(h=0, lwd=1, col="black", lty="dotted")

modelout_mod_R18 = modelout_mod_R18[order(modelout_mod_R18$mean_hb), , drop = FALSE]
plot(1:nrow(modelout_mod_R18), modelout_mod_R18$mean_hb, las = 1, pch=19, col="#c96d00", ylim=c(-.75,.75), ylab="honey bee slope (mean and 95% CI)", cex=1.5, xaxt = "n", xlab="crop study", main="Honey Bee")
axis(1, at=seq(0,nrow(modelout_mod_R18),10), labels=seq(0,nrow(modelout_mod_R18),10))
arrows(1:nrow(modelout_mod_R18), modelout_mod_R18$ci_lower_hb, 1:nrow(modelout_mod_R18), modelout_mod_R18$ci_upper_hb, length=0.05, angle=90, code=3, col="#c96d00", lwd=1.5)
abline(h=a_hb, lty="dashed", lwd=2)
abline(h=0, lwd=1, col="black", lty="dotted")

modelout_mod_R18 = modelout_mod_R18[order(modelout_mod_R18$mean_rich), , drop = FALSE]
plot(1:nrow(modelout_mod_R18), modelout_mod_R18$mean_rich, las = 1, pch=19, col="#00a2ca", ylim=c(-.75,.75), ylab="richness slope (mean and 95% CI)", cex=1.5, xaxt = "n", xlab="crop study", main="Richness")
axis(1, at=seq(0,nrow(modelout_mod_R18),10), labels=seq(0,nrow(modelout_mod_R18),10))
arrows(1:nrow(modelout_mod_R18), modelout_mod_R18$ci_lower_rich, 1:nrow(modelout_mod_R18), modelout_mod_R18$ci_upper_rich, length=0.05, angle=90, code=3, col="#00a2ca", lwd=1.5)
abline(h=a_rich, lty="dashed", lwd=2)
abline(h=0, lwd=1, col="black", lty="dotted")

```

**Fig 2:** This figure shows the estimated means and 95% CIs for the effect of wild insects, honey bee, and species richness on crop yield for each pollination study. The dashed black line is the overall mean across all studies. Means were calculated using the random effect estimates from the full model.

### 3) Is the total number of flower visits by pollinators sufficient to predict crop yields, or is the diversity of pollinator species also important?

The mean wild insects estimate is `r round(a["wild_insects_z"], 3)`, and the honeybee mean estimate is `r round(a["honeybee_z"], 3)`. Overall, species richness has a shallower positive effect on yield (pollinator richness estimate is `r round(a["richness_z"],3)`)

```{r Fig3, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}


# plot point estimates with CIs
par(mfrow=c(1,1))
par(mar = c(3,3,3,3))

# plot coefficients from Model 5 (89 studies)
thisout = data.frame(par = c("Wild insects", "Honey bee", "Richness"), 
	mean_modR18 = c(summary(mod_R18)$coefficients[2], summary(mod_R18)$coefficients[3], summary(mod_R18)$coefficients[4]),
	CIlower_mod_R18 = c(summary(mod_R18)$coefficients[2] - 1.96*summary(mod_R18)$coefficients[9], summary(mod_R18)$coefficients[3] - 1.96*summary(mod_R18)$coefficients[10], summary(mod_R18)$coefficients[4] - 1.96*summary(mod_R18)$coefficients[11]),
	CIuppermod_R18 = c(summary(mod_R18)$coefficients[2] + 1.96*summary(mod_R18)$coefficients[9],summary(mod_R18)$coefficients[3] + 1.96*summary(mod_R18)$coefficients[10], summary(mod_R18)$coefficients[4] + 1.96*summary(mod_R18)$coefficients[11])
	)
	
plot(1:3, thisout$mean_modR18, pch=19, col=c("#00bb5c", "#c96d00","#00a2ca"), cex=1.75, las = 1, ylim=c(-0.1,0.3),  xlab = NA, xaxt = 'n', xlim=c(.5,3.5), ylab="Overall slope (mean, 95% CI)", main="")
axis(1, at=1:3, labels=thisout$par, srt=90)

abline(h=0, lty="dashed", col="gray")


arrows(
	1:3, thisout$CIlower_mod_R18,
	1:3, thisout$CIuppermod_R18,
       length=0.1, angle=90, code=3, col=c("#00bb5c", "#c96d00","#00a2ca"), lwd=2.5)

```

**Fig 3:** This figure shows the current overall mean estimate (and 95% CIs) for the effect of wild insects, honey bees, and richness on crop yield. Estimates based on the CropPol database are generated using the full model as described in the methods.

### 4) How the relationships between pollinator visits, richness, and yield changed with the number of studies available?

Since the first synthesis papers, we found that the effect size tend to decrease as more studies are added to the analysis.

```{r Fig4, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

#Data is taken from estimates_data.R

estimates <- read.csv("scripts/estimates_data.csv")

par(mfrow=c(1,1))
par(mar = c(5,3,3,3))
plot(1:nrow(estimates),estimates$estimate_HB, col = "white", las = 1, xlab = "", ylab = "estimates", ylim = c(-0.1, 0.4),  xaxt = "n")
axis(1, at = 1:nrow(estimates), labels = estimates$timepoint, las=2, cex.axis = 0.7)

points(which(!is.na(estimates$estimate_HB)),subset(estimates, !is.na(estimate_HB))$estimate_HB, t = "b", col = "#00bb5c", lwd=2)
points(which(!is.na(estimates$estimate_WI)),subset(estimates, !is.na(estimate_WI))$estimate_WI, t = "b", col = "#c96d00", lwd=2)
points(which(!is.na(estimates$estimate_richness)),subset(estimates, !is.na(estimate_richness))$estimate_richness, t = "b", col = "#00a2ca", lwd=2)
abline(h=0, lty="dashed", col="gray")
text(y = rep(-0.075, nrow(estimates)), 
     x = 1:nrow(estimates),
     labels = paste0("(",estimates$n, ")"), col = "gray", cex = 0.5)
      

legend("topright", inset=.03, legend=c("wild insects", "honey bee", "richness"), col=c("#c96d00", "#00bb5c", "#00a2ca"), lty=1, lwd=2, box.lty=0, cex = 0.7)
# Maybe we can add a legend and use somehow the info on sample size.

```

**Fig 4:** This figure shows the history of the estimated effects of wild insects, honey bees, and richness on crop yield over time. Literature estimates are drawn from Garibaldi et al. 2013, Garibaldi et al. 2015, Rader et al. 2016, and Dainesse et al. 2019. Estimates based on the CropPol database are generated using the full model as described in methods. Sample size is indicated between brackets below each time period.
